### 2015-07-29 doc/thesis/system.tex
Several fixes to the text about typing rules

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{\overline{F{:}V}\}_{unique|open|fixed|closed}$,
where the notation $\overline{F{:}V}$ denotes the list $F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}$.
Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
Value types represent mutable values by default, but we can use the
$\Const$ type to make them represent immutable values.

\label{def:tabletype}
We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
to classify table types.
The tag \emph{unique} represents tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{fixed} represents \emph{unique} table types that we
do not know how many aliases they have.
In particular, the type of a class has this tag.
The tag \emph{closed} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

@@

With the recursive types we finish the dicussion about Typed Lua types,
and we begin the discussion about the special types.

@@

\label{def:filtertype}
Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we mentioned in Section \ref{sec:unions}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

@@

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} to type check
the right-hand side of the assignment.
The rule \textsc{T-EXPLIST1} is the rule that type checks an
expression list where all expressions can only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{f}}
\end{array}
\]

@@

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
check a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{f}}
\end{array}
\]

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that expresses this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:E_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:E_{2}, \env_{3} \;\;\;
       E_{1} \lesssim E_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

@@

Rules for function applications are similar to the rule for multiple assignment.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:E_{1} \rightarrow E_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:E_{3}, \env_{3} \;\;\;
       E_{3} \lesssim E_{1}}
      {\env_{1}, \penv \vdash e(el):E_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
that produce no value or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(E), \env_{2}}
\end{array}
\end{array}
\]

We can define \emph{first} inductively as follows:
\begin{align*}
first(F{*}) & = F \cup \Nil\\
first(F \times P) & = F\\
first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
\end{align*}

@@

First, Typed Lua uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(F,V)$ and a new type environment $\env_{2}$.
A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
where $F$ is the type of the key and $V$ is the type of the value.

After type checking each table field, our type system uses each pair $(F,V)$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\begin{array}{rcl}
wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
wf({\bf const}\; F) & = & wf(F) \\
wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
wf(\mu x.F) & = & wf(F) \\
wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
wf(F{*}) &=& wf(F) \\
wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
tag(\{\overline{F:V}\}_t, t) & = & \top\\
\end{array}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
Due to this restriction, the rule that type check table fields
use the auxiliary function \emph{fix} (in the definition of \emph{vt})
to change any \emph{unique} and \emph{open} table types
used in the field initializer to \emph{fixed}.
Rule \textsc{T-FIELD} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-FIELD}\\
\dfrac{\env_{1}, \penv \vdash e_{2}:V, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{1}:F, \env_{3}}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (F,vt(F,V)), \env_{3}}
\end{array}
\]

The rule \textsc{T-FIELD} uses the auxiliary function \emph{vt}
to adjust the type of the value according to the type of the key.
More precisely, \emph{vt} includes the type $\Nil$ in the type of
the value when the type of the key is not a literal type.
It also uses \emph{fix} to prevent \emph{unique} and \emph{open}
table types to appear in the type of the values.
We define these functions as follows:
\begin{align*}
vt(L, V) & = fix(V) \\
vt(F_1, F_2) & = nil(fix(F_2))\\
vt(F_1, {\bf const}\;F_2) & = {\bf const}\;nil(fix(F_2))\\
\\
nil(T) & = T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
nil(T) & = T \cup \Nil \;\;\; \mathrm{otherwise}\\
\\
fix(F_{1} \cup F_{2}) & = fix(F_{1}) \cup fix(F_{2})\\
fix(\{\overline{F{:}V}\}_{unique|open}) & = \{\overline{F{:}V}\}_{fixed} \\
fix(F) & = F
\end{align*}

The rule \textsc{T-CONSTRUCTOR1} uses these steps to type check a
table constructor with a list of table fields that do not end with
an expression that potentially returns multiple values:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \\
       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

@@

We use the refinement of table types to handle the declaration of
new global variables.
In Lua, the assignment \texttt{v = v + 1} translates to
\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an \emph{open} table in the top-level scope.
In the following examples we assume that $\string_ENV$ is in the
environment and has type $\{\}_{open}$.

@@

The rules for declaring variadic functions and recursive functions
are similar to \textsc{T-FUNCTION1}, and we will not
show them in this section for brevity.

@@

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
The function \emph{idiv} has type
\[
\Integer \times \Integer \times \Nil{*} \rightarrow (\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
integer division.
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that describes
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Nil{*}$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : E, \env_{2}, (x,S)$
for type checking $idiv(1, 2)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $E$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S)$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S$ that should be bound
to a fresh variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST4} for
type checking $idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\;
       n = |\;\overline{e}\;| \\
       S = T_{n+1} \times ... \times T_{n+m} \times \Nil{*} \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Nil{*} \;\;\;
       x \notin \penv \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2})
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
\end{array}
\]

Note that $idiv(1, 2)$ has type
$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
\[
(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})) 
\]

@@

Typed Lua does not allow assignments to left-hand values that are bound
to a projection type.
This kind of assignment would be unsound, because it could break the
dependency relation that the components of each tuple of the union have.
For instance, the following example does not type check:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$r = ``foo";$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

>>>>

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{\overline{F{:}V}\}_{unique|open|fixed|closed}$,
where the notation $\overline{F{:}V}$ denotes the list $F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}$.
Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
Value types represent mutable fields by default, but we can use the
$\Const$ type to make them represent immutable fields.
Making a field $\Const$ does not guarantee that its value cannot change,
as the table may have aliases with a non-$\Const$ type for that field.
Typed Lua needs immutable fields to enable depth subtyping between table types.

\label{def:tabletype}
We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
to classify table types.
The tag \emph{closed} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{closed} table types.
The tags \emph{unique}, \emph{open}, and \emph{fixed} represent
tables with no keys that do not inhabit one of the table's key types,
but with different guarantees about the reference to a value of
that type.
A reference to an \emph{unique} table is guaranteed to point
to a table that has no other references to it.
In particular, the type of the table constructor has this tag,
as it allows greater flexibility in reinterpreting its type.
A reference to an \emph{open} table is guaranteed to have only
\emph{closed} references pointing to the same table,
a guarantee that still lets the type system reinterpret the type
of a reference, but with more restrictions.
A reference to a \emph{fixed} table can have any number of
\emph{fixed} or \emph{closed} references point to it,
so its type cannot change anymore.
In particular, the type of a class has this tag in our type system.

A \emph{fixed} table type guarantees that there are
no keys with a type that is not one of its key types.
Even though this guarantee allows type-safe iteration on
\emph{fixed} table types, it forbids width subtyping that
is necessary for object-oriented programming, so \emph{closed}
table types remove this guarantee to allow width subytping
between other table types and \emph{closed} table types.
This means that objects have \emph{closed} table types,
while their classes have \emph{fixed} table types.

@@

With recursive types we finish the dicussion about Typed Lua types,
and we begin the discussion about special types.

@@

\label{def:filtertype}
Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we mentioned in Section \ref{sec:unions}.
This means that \texttt{type} predicates use filter types of the form
$\phi(F_{1},F_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(F_{1},T_{2})$, $F_{1}$ is the original type and
$F_{2}$ is the discriminated type.

@@

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} to type check
the right-hand side of the assignment.
The rule \textsc{T-EXPLIST1} is the rule that type checks an
expression list where all expressions can only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{1}, \penv \vdash e_{i}:F_{i}, \env_{i+1} \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:F_{1} \times ... \times F_{n} \times \Nil{*}, \env_{f}}
\end{array}
\]

@@

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
check a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:F_{i}, \env_{i+1} \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:F_{1} \times ... \times F_{n} \times \Value{*}, \env_{f}}
\end{array}
\]

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that expresses this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

@@

Rules for function applications are similar to the rule for multiple assignment.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
       S_{3} \lesssim S_{1}}
      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
that produce no value or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{proj} to ensure that only the first value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:proj(S,1), \env_{2}}
\end{array}
\end{array}
\]

We can define \emph{proj} inductively as follows:
\begin{align*}
proj(S_1 \sqcup S_2, i) & = proj(S_1, i) \cup proj(S_2, i) \\
proj(F{*}, i) & = nil(F) \\
proj(F \times P, 1) & = F \\
proj(F \times P, i) & = proj(P, i-1)\\
proj(E{*}, i) & = nil(E) \\
proj(T \times E, 1) & = T \\
proj(T \times E, i) & = proj(E, i-1) \\
nil(T) & = \left\{
\begin{array}{ll}
T & \text{if $\Nil \lesssim T$} \\
T \cup \Nil & \text{otherwise}
\end{array} \right.
\end{align*}

@@

First, Typed Lua uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(F,V)$ and a new type environment $\env_{2}$.
A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
where $F$ is the type of the key and $V$ is the type of the field value.

After type checking each table field, our type system uses each pair $(F,V)$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
The predicate \emph{wf} also uses the auxiliary predicate \emph{tag} to
forbid \emph{unique} and \emph{open} fields.
Formally, a table type is well-formed if it obeys the following rule:
\[
\begin{array}{rcl}
wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
wf({\bf const}\; F) & = & wf(F) \\
wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
wf(\mu x.F) & = & wf(F) \\
wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
wf(F{*}) &=& wf(F) \\
wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}
\end{array}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the field values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
Due to this restriction, the rule that type check table fields
use the auxiliary function \emph{fix} (in the definition of \emph{vt})
to change any \emph{unique} and \emph{open} table types
used in the field initializer to \emph{fixed}.
Rule \textsc{T-FIELD} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-FIELD}\\
\dfrac{\env_{1}, \penv \vdash e_{2}:V, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{1}:F, \env_{3}}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (F,vt(F,V)), \env_{3}}
\end{array}
\]

The rule \textsc{T-FIELD} uses the auxiliary function \emph{vt}
to adjust the type of the value according to the type of the key.
More precisely, \emph{vt} includes the type $\Nil$ in the type of
the value when the type of the key is not a literal type.
It also uses \emph{fix} to prevent \emph{unique} and \emph{open}
table types to appear in the type of the field values.
We define these functions as follows:
\begin{align*}
vt(L, V) & = fix(V) \\
vt(F_1, F_2) & = nil(fix(F_2))\\
vt(F_1, {\bf const}\;F_2) & = {\bf const}\;nil(fix(F_2))\\
\\
fix(F_{1} \cup F_{2}) & = fix(F_{1}) \cup fix(F_{2})\\
fix(\{\overline{F{:}V}\}_{unique|open}) & = \{\overline{F{:}V}\}_{fixed} \\
fix(F) & = F
\end{align*}

The rule \textsc{T-CONSTRUCTOR1} uses these steps to type check a
table constructor with a list of table fields that do not end with
an expression that potentially returns multiple values:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \;\;\;
       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \\
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

@@

We use the refinement of table types to handle the declaration of
new global variables.
In Lua, the assignment \texttt{v = v + 1} translates to the statement
\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an \emph{open} table in the top-level scope.
In the following examples we assume that $\string_ENV$ is in the
environment and has type $\{\}_{open}$.

@@

The rules for declaring variadic functions and recursive functions
are similar to \textsc{T-FUNCTION1}, and we did not discuss them in
this section for brevity.

@@

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
integer division and has type
\[
\Integer \times \Integer \times \Nil{*} \rightarrow (\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that describes
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Nil{*}$.
Let us also assume that $a$ and $b$ are local variables in the environment,
and that both have type $\Integer$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(a, b) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : E, \env_{2}, (x,S)$
for type checking $idiv(a, b)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $E$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S)$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S$ that should be bound
to a fresh variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST3} for
type checking $idiv(a, b)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:F_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
       S = F_{n+1} \times ... \times F_{n+m} \times \Nil{*} \sqcup F_{n+1}' \times ... \times F_{n+m}' \times \Nil{*} \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:F_{1} \times ... \times F_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
\end{array}
\]

Note that $idiv(a, b)$ has type
$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
\[
(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})) 
\]

@@

Typed Lua does not allow assignments to left-hand values that are bound
to a projection type.
This kind of assignment would be unsound, because it could break the
dependency relation that the components of each tuple of the union have.
For instance, the following example does not type check:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(a, b) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$r = ``foo";$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}
